Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 8: make a small note.
Sections 1-3: Fiddling.

This report attempts to summarize the various aspects of the workshop.
The remainder of this paper first gives an overview of the process
with which the workshop was put together (\S\ref{sec:process}), then
proceeds with summaries of the two keynote
presentations~(\S\ref{sec:keynotes}), followed by summaries of the
workshop papers grouped by the three thematic workshop panels to which
they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Two
broader cross-cutting issues surfaced repeatedly, and are discussed
separately~(\S\ref{sec:cross-cutting}), as are use cases for
sustainable software~(\S\ref{sec:use-cases}). The summaries are based
not only on the papers and panel presentations, but also on the many
comments raised in both the onsite and online discussions that
accompanied the workshop elements, as documented by collaborative note
taking during the workshop~\cite{WSSSPE1-google-notes}. We conclude
with issues and lessons learned, as well as plans for future
activities~(\S\ref{sec:conclusions}).  The short papers accepted to
the workshop are listed in Appendix~\ref{sec:papers}, and a partial
list of workshop attendees can be found in
Appendix~\ref{sec:attendees}.

@@

The WSSSPE1 workshop began with two keynote presentations.

@@

Another problem Bourne experienced was staff retention from private
organizations which reward those combining research and software
expertise (the ``Google Bus''). However, he is a strong fan of
software sustainability through public-private partnerships. He noted
that making a successful business from scientific software alone is
rare: founders overvalue while customers undervalue. He noted that to
last, an open source project needs a minimal funding requirement even
with a vibrant community---goodwill only goes so far if one is being
paid to do something else.  He talked about grant schemes of relevance
in the U.S., particularly with regard to technology
transfer~\cite{sbir-web, fased-web}.

@@

The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon
Smith of GitHub. Smith started with an example from his past in data
reduction in Astronomy, where he needed to remove interfering effects
from the experimental apparatuses. He built a ``bad pixel mask,'' and
realized that while it was persistent, there was no way or practice of
sharing these data among scientists. Consequently many researchers
repeated the same calculations. Smith estimated that 13 person-years
were wasted by this repetition.

@@

C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor, argued that with
regard to reusable software, ``we should just start doing it.''  Smith
replied that documentation should be ``treated as a first class
entity.''  He noted that the open source community has excellent
cultures of code reuse, for example, RubyGems~\cite{rubygems-web},
PyPI~\cite{pypi-web}, and CPAN~\cite{cpan-web}, where there is
effectively low-friction collaboration through the use of
repositories. This has not happened in highly numerical, compiled
language scientific software.  An exception he cited as a good example
of scientific projects using GitHub is the EMCEE Markov Chain Monte
Carlo project~\cite{emcee-web} by Dan Foreman-Mackey and contributors.

@@

% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of
software projects wildly vary in factors such as scope, purpose and
usage. The development and management of each kind is significantly
different. Consequently, the sustainability challenges faced by them
differ and must be treated separately. For instance, the challenges
faced by a gateway software development project such as
\toolname{CIPRES} or visualization software pinroducts such as
\toolname{VisIT} or \toolname{ParaView} are distinct to a niche software for
\textit{ab initio} modeling and simulation such as \toolname{VASP} or
\toolname{Quantum Espresso}. \todo{cite}

>>>>

This report attempts to summarize the various aspects of the workshop.
The remainder of this paper first gives an overview of the process
with which the workshop was put together (\S\ref{sec:process}), then
proceeds with summaries of the two keynote
presentations~(\S\ref{sec:keynotes}), followed by summaries of the
workshop papers grouped by the three thematic workshop panels to which
they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Two
broader cross-cutting issues surfaced repeatedly, and are discussed
separately~(\S\ref{sec:cross-cutting}), as are use cases for
sustainable software~(\S\ref{sec:use-cases}). The summaries are based
not only on the papers and panel presentations, but also on the many
comments raised in both the onsite and online discussions that
accompanied the workshop elements, as documented by collaborative note
taking during the workshop~\cite{WSSSPE1-google-notes}. We conclude
with issues and lessons learned, as well as plans for future
activities~(\S\ref{sec:conclusions}).  The short papers accepted to
the workshop are listed in Appendix~\ref{sec:papers}, and a partial
list of workshop attendees can be found in
Appendix~\ref{sec:attendees}. The original call for papers is included in 
Appendix~\ref{sec:cfp}.

@@

The WSSSPE1 workshop began with two keynote presentations, which 
resonated with the audience and spurred a number of topics  
discussed throughout the meeting.

@@

Another problem Bourne experienced was staff retention from private
organizations which reward those combining research and software
expertise (the ``Google Bus''). However, he is a strong supporter of
software sustainability through public-private partnerships. He noted
that making a successful business from scientific software alone is
rare: founders overvalue while customers undervalue. He noted that to
last, an open source project needs a minimal funding requirement even
with a vibrant community --- goodwill only goes so far if one is being
paid to do something else.  He talked about grant schemes of relevance
in the U.S., particularly with regard to technology
transfer~\cite{sbir-web, fased-web}.

@@

The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon
Smith of GitHub. Smith started with an example from his past in data
reduction in Astronomy, where he needed to remove interfering effects
from the experimental apparatuses. He built a ``bad pixel mask,'' and
realized that while it was persistent, there was no way or practice of
sharing these data among scientists. Consequently many researchers
repeated the same calculations. Smith estimated that plausibly 13 person-years
were wasted by this repetition.

@@

C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor, argued that with
regard to reusable software, ``we should just start doing it.''  Smith
replied that documentation should be ``treated as a first class
entity.''  He noted that the open source community has excellent
cultures of code reuse, for example, RubyGems~\cite{rubygems-web},
PyPI~\cite{pypi-web}, and CPAN~\cite{cpan-web}, where there is
effectively low-friction collaboration through the use of
repositories. This has not happened in highly numerical, compiled
language scientific software.  An exception he cited as a good example
of scientific projects using GitHub is the EMCEE Markov Chain Monte
Carlo project~\cite{emcee-web} developed by Dan Foreman-Mackey and contributors.

@@

% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of
software projects wildly vary in factors such as scope, purpose and
usage. The development and management of each kind is significantly
different. Consequently, the sustainability challenges faced by them
differ and must be treated separately. For instance, the challenges
faced by a gateway software development project such as
\toolname{CIPRES} or visualization software pinroducts such as
\toolname{VisIT} or \toolname{ParaView} are distinct to a niche software for
\textit{ab initio} modeling and simulation such as \toolname{VASP} or
\toolname{Quantum Espresso}. \note{cite SW}
